---
title: "Getting Started with mainger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mainger}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5
)
set.seed(2026)
```

## Overview

The `mainger` package provides a unified interface for integrating external
summary statistics into an internal regression model under varying data-sharing
constraints. The package implements the Mahalanobis-integrated estimator and
its closed-form variants across three sharing regimes:

| Regime       | Internal access            | External access                         |
|--------------|----------------------------|------------------------------------------|
| Full         | $(X_1, Y_1)$               | $\hat\theta$ and $\hat\Sigma_2$          |
| Partial      | $(X_1, Y_1)$               | $\hat\theta$ only                        |
| Restricted   | $r_1 = X_1^\top Y_1 / n_1$ | $\hat\theta$ and a reference $\Sigma_{\mathrm{ref}}$ |

The same top-level function `mainger()` handles all three; the regime is inferred
from which inputs you supply. The package then computes a theoretical bound on
the tuning parameter $\eta$, selects $\eta$ by cross-validation (full and partial)
or extended AIC (restricted), and returns an integrated coefficient vector with
diagnostics.

This vignette walks through one example per regime on simulated data.

```{r load}
library(mainger)
```

## A common data-generating setup

We use the same internal design across all three examples so the regimes can be
compared at fixed sample size and noise level. The internal cohort is small
(`n_int = 80`) with `p = 20` predictors and a sparse true coefficient
(five active variables).

```{r dgp}
n_int <- 80; n_ext <- 5000; p <- 20

X_int <- matrix(rnorm(n_int * p), n_int, p)
colnames(X_int) <- paste0("X", seq_len(p))
beta_true <- c(rep(0.5, 5), rep(0, p - 5))
Y_int <- as.numeric(X_int %*% beta_true + rnorm(n_int, 0, 1))

# External cohort with mild coefficient heterogeneity
beta_ext_true <- beta_true + rnorm(p, 0, 0.05)
X_ext <- matrix(rnorm(n_ext * p), n_ext, p)
Y_ext <- as.numeric(X_ext %*% beta_ext_true + rnorm(n_ext, 0, 1))

# External summary: coefficient and Gram matrix
theta_ext <- as.numeric(solve(crossprod(X_ext), crossprod(X_ext, Y_ext)))
Sigma_ext <- crossprod(X_ext) / n_ext
```

## Example 1: Partial sharing

The most common scenario in clinical research: the external collaborator
shares a fitted coefficient vector but not their Gram matrix. Pass `X_int`,
`Y_int`, and `beta_ext`:

```{r partial-fit}
fit_partial <- mainger(X_int = X_int, Y_int = Y_int, beta_ext = theta_ext)
fit_partial
```

The package detects the **partial** regime automatically (because no external
Gram matrix was supplied), computes the theoretical bound on $\eta$, builds
a search grid bounded by `1.5 * eta_bound`, and selects $\eta$ via 5-fold
cross-validation.

```{r partial-summary}
summary(fit_partial)
```

Inspect the integrated coefficients and compare against the internal-only
baseline:

```{r partial-coef}
internal_only <- as.numeric(solve(crossprod(X_int), crossprod(X_int, Y_int)))

data.frame(
  variable   = colnames(X_int),
  truth      = beta_true,
  internal   = round(internal_only, 3),
  external   = round(theta_ext, 3),
  integrated = round(fit_partial$coefficients, 3)
)[1:10, ]
```

A held-out test set lets us verify the integrated estimator has lower
prediction error than the internal-only fit:

```{r partial-test}
X_test <- matrix(rnorm(2000 * p), 2000, p)
Y_test <- as.numeric(X_test %*% beta_true + rnorm(2000, 0, 1))

mspe <- function(beta) mean((Y_test - X_test %*% beta)^2)
c(internal   = mspe(internal_only),
  external   = mspe(theta_ext),
  integrated = mspe(fit_partial$coefficients))
```

## Example 2: Full sharing

If the external Gram matrix is also available, the integrated estimator
uses a matrix-weighted combination rather than a scalar-weighted one. Pass
`Sigma_ext` in addition:

```{r full-fit}
fit_full <- mainger(X_int = X_int, Y_int = Y_int,
                    beta_ext = theta_ext, Sigma_ext = Sigma_ext,
                    sigma2_ext = 1, n_ext = n_ext)
fit_full
```

The full regime in addition to selecting $\eta$ reports a **concordance**
check: whether the eigenstructure of $\Sigma_{\mathrm{int}}^{-1}\Sigma_{\mathrm{ext}}$
favors full sharing or partial sharing at the selected $\eta$. The check is
the empirical realization of Proposition 2 in the accompanying paper.

```{r full-concordance}
fit_full$concordance
```

A negative `advantage_at_eta` means partial sharing outperforms full sharing
at the selected $\eta$; positive means full sharing wins. The package will
recommend the better regime automatically when they disagree:

```{r full-diagnose, eval = FALSE}
diagnose(fit_full)
```

## Example 3: Restricted sharing

When individual-level internal data cannot be shared (genomics, federated
analysis), only marginal correlations $r_1 = X_1^\top Y_1 / n_1$ and a
reference covariance $\Sigma_{\mathrm{ref}}$ are available. Pass these
instead of `X_int` and `Y_int`:

```{r restricted-setup}
# Simulate the restricted-access scenario from the same internal cohort
r_int     <- as.numeric(crossprod(X_int, Y_int)) / n_int
Sigma_ref <- crossprod(X_int) / n_int  # Reference panel; could be a public panel
```

```{r restricted-fit}
fit_restricted <- mainger(r_int = r_int, Sigma_ref = Sigma_ref,
                          beta_ext = theta_ext,
                          n_int = n_int, sigma2_int = 1)
fit_restricted
```

The restricted regime uses extended AIC instead of cross-validation because
held-out folds cannot be constructed without individual data.

## Diagnostics and the tuning grid

For any regime, the diagnostics surface the search behavior:

```{r diagnostics}
fit_partial$eta            # Selected eta
fit_partial$eta_bound      # Theoretical upper bound used for the grid
fit_partial$eta_grid       # Full search grid
```

The selected $\eta$ is bounded above by `1.5 * eta_bound` in normal cases;
when the bound is degenerate (zero or undefined), the package falls back
to a fixed `[0, 5]` grid. See `?eta_bound_partial` for the bound formula.

## S3 methods

Standard `predict`, `summary`, `plot`, and `coef` methods are available:

```{r methods, eval = FALSE}
predict(fit_partial, newdata = X_test)
plot(fit_partial)                          # eta-vs-objective curve
plot(fit_full,    type = "advantage")      # Full-vs-Partial concordance plot
coef(fit_restricted)
```

## Reproducibility

All three examples above use `cv_seed = NULL` (random folds). To produce
bit-identical results across runs, pass `cv_seed`:

```{r cvseed, eval = FALSE}
fit_partial <- mainger(X_int = X_int, Y_int = Y_int, beta_ext = theta_ext,
                       tuning = "cv", cv_seed = 548)
```

The same seed is used throughout the package for cross-validation, allowing
the integration result to be reproduced across machines.

## Next steps

- `?mainger` documents every argument and return field.
- `?eta_bound_partial` and `?eta_bound_full` describe the theoretical
  bounds used to construct the search grid.
- `?check_concordance` and `?spectral_advantage` describe the diagnostics
  used in the full-sharing regime.
